---
title: "Vix and Future Volatility"
output: html_notebook
---
 
```{r setup, include = FALSE}
library(tidyquant)
library(tidyverse)
```

In a [previous post](http://www.reproduciblefinance.com/2017/08/28/realized-volatility-and-the-vix/), we examined the relationship between the [VIX](http://www.cboe.com/products/vix-index-volatility/vix-options-and-futures/vix-index) and the past, realized volatility of the S&P 500. 

Today, we'll think about how/whether the VIX predicts future volatility and we'll be reproducing some visualizations from [this Bloomberg piece](https://www.bloomberg.com/view/articles/2017-07-03/what-history-says-about-low-volatility).  

Once again, the substance and ideas here are 100% attributable to Bloomberg - my goal is to reproduce and add to our R toolkit, and learn something about volatility.

First, we grab the price history of the VIX and SP500, convert to returns, and calculate the rolling 20-day volatility of the SP500. 

```{r, message=FALSE, warning=FALSE}
symbols <- c("^GSPC", "^VIX")

prices <- 
  getSymbols(symbols, src = 'yahoo', from = "1990-01-01", 
             auto.assign = TRUE, warnings = FALSE) %>% 
  map(~Cl(get(.)))
  
# get daily, cont compounded returns
sp500_returns <- na.omit(ROC(GSPC$GSPC.Close, 1, type = "continuous"))

  
sp500_rolling_sd_20 <- rollapply(sp500_returns,
                             20,
                             function(x) StdDev(x))
  
sp500_rolling_sd_annualized_20 <- (round((sqrt(252) * sp500_rolling_sd_20 * 100), 2))

vol_vix_xts <- merge.xts(VIX$VIX.Close, sp500_rolling_sd_annualized_20)
```

We have VIX prices and rolling 20-day SP500 volatility. We want to reproduce the visualizations that illuminate how the VIX predicts realized volatility. In other words, is the VIX reading on 2017-01-01 telling us anything aobut realized volatility 20 days from 2017-01-01? 

The data wrangling challenge here is all about the dates. We need to compare the VIX reading on 2017-01-01 to the realized volatility that occurs over the 20 days after 2017-01-01. It's not hard to understand this in our heads, it's a little tricky to line the data up correctly (full disclosure: I had to go to a whiteboard and draw out 20 cells with arrows for what need to go where from the data frame.)

First, we'll set a start date as the first date index of `sp500_rolling_sd_annualized_20` and an end date for the date index that is 19 days before end of `sp500_rolling_sd_annualized_20`.

```{r, message = FALSE, warning = FALSE}

start.date <- index(sp500_rolling_sd_annualized_20[1])

end.date <- index(sp500_rolling_sd_annualized_20[nrow(sp500_rolling_sd_annualized_20) - 19])

```

Ok, we will start on `r start.date` and end on `r end.date`.

Now let's grab the SP500 rolling volatility data between those dates. Note we use `coredata()` function because we don't need the time index, just the data.

```{r, message = FALSE, warning = FALSE}
# Extract the data, not the time index since we're adding to a tibble.
future_sp500_vol <- as_tibble(coredata(sp500_rolling_sd_annualized_20[20:nrow(sp500_rolling_sd_annualized_20), ]))
```

Now we want to get the VIX data for those same dates and combine them into one tibble. We will get the VIX data with `VIX$VIX.Close[paste(start.date, end.date, sep = "::")]`.

```{r, message = FALSE, warning = FALSE}

vix_sp500_future_vol <- 
  # Get VIX data
  VIX$VIX.Close[paste(start.date, end.date, sep = "::")] %>%
  # Add a date column.
  as_tibble(preserve_row_names = TRUE) %>% 
  mutate(date = ymd(row.names)) %>%
  # Delete row.names.
  select(-row.names) %>% 
  select(date, everything()) %>%
  # Add the SP500 data with bind_cols - important line of code here!
  bind_cols(future_sp500_vol[1]) %>% 
  rename(vix = VIX.Close, Spy.future = GSPC.Close) 
 
tail(vix_sp500_future_vol)
```

Take a peek at the result. We now have a tibble with a column for the VIX price on date t, and a column for realized, future volatility 20 days after t.

Now it's on to `ggplot` to recreate the visualizations from Bloomberg.

```{r}
line_future_vol20 <- 
  # Plot vix on the x axis and future vol on the y axis
  ggplot(vix_sp500_future_vol, aes(date)) + 
  geom_line(aes(y = Spy.future, colour = "SP500")) +
  geom_line(aes(y = vix, colour = "Vix")) +
  scale_color_manual(values = c(SP500 = 'maroon', Vix = 'cornflowerblue')) +
  # The remaining lines are aesthetic, title, axis labels.
  ggtitle("Vix and SP500 Volatility 20 Trading Days Hence") +
  xlab("date") +
  ylab("percent") +
  scale_y_continuous(labels = function(x){ paste0(x, "%") }, 
                     breaks = scales::pretty_breaks(n=10)) +
  theme(plot.title = element_text(hjust = 0.5))

line_future_vol20
```

I can't eyeball much from that chart, let's go to a scatter plot. Before we do so, let's use nested if/else statements to create three eras: pre-historic, Obama, Trump,


```{r, message=FALSE, warning=FALSE}
vix_sp500_future_vol$era <- 
  ifelse(vix_sp500_future_vol$date <= "2008-11-03", 'pre-historic',
         ifelse(vix_sp500_future_vol$date >"2008-11-04" & 
                  vix_sp500_future_vol$date <=  "2016-11-08", 'Obama', 'Trump'))
```



```{r, message=FALSE, warning=FALSE}
scatter_future_vol20 <- 
  ggplot(vix_sp500_future_vol, aes(vix, Spy.future, color = era, shape = era)) + 
  geom_point() +
  scale_color_manual(values=c("maroon", "cornflowerblue", "green")) +
  scale_shape_manual(values = c(17, 15, 18)) +
  geom_line(aes(vix, vix), color = "black") +
  # The remaining lines are aesthetic, title, axis labels.
  ggtitle("Vix v. Subsequent Volatility") +
  xlab("Vix") +
  ylab("Realized vol 20 subsequent trading days") +
  scale_y_continuous(labels = function(x){ paste0(x, "%") }, 
                     breaks = scales::pretty_breaks(n=10)) +
  scale_x_continuous(labels = function(x){ paste0(x, "%") },
                     breaks = scales::pretty_breaks(n=10)) +
  theme(plot.title = element_text(hjust = 0.5))

scatter_future_vol20
```

That diagonal black line is the volatility predicted by the VIX and we can see that actual volatility tends to be a bit lower than predicted. The maroon diamonds are the observations durning the Obama administration. The green triangles are the observations since November 8, 2016. Over the last ~9 years, realized volatility has generally been lower than what the VIX was predicting.

For the well-articulated insights and theory, have a look at the original [piece on Bloomberg](https://www.bloomberg.com/view/articles/2017-07-03/what-history-says-about-low-volatility). 

Another interesting note is that since 1990, we have never seen subsequent one-month volatiliy reach 20% when the VIX was below 12% (add a vertical line at 12% to see this more clearly). Said another way, spikes in realized volatility tend to follow spikes in the VIX.   

That's all for today. The main goal was to reproduce visualizations and insights about the VIX and realized volatility.  Thanks for reading and next time we'll wrap our VIX work into a Shiny app.
