---
title: "Vix and Future Volatility"
output: html_notebook
---
 
```{r setup, include = FALSE}
library(tidyquant)
library(tidyverse)
library(quantmod)
library(broom)
```

In our last post, we examined the relationship between the [VIX](http://www.cboe.com/products/vix-index-volatility/vix-options-and-futures/vix-index) and the past, realized volatility of the S&P 500. 

The VIX is a measure of the expected future volatility of the S&P500 and it has been [quite low recently](https://www.cnbc.com/2017/07/26/wall-streets-fear-index-the-vix-falls-to-record-low-under-9-after-fed-statement.html).  As a volatility nerd, I came across an interesting piece from [Bloomber on the meaning of the VIX](https://www.bloomberg.com/view/articles/2017-07-03/what-history-says-about-low-volatility). As a reproducibility and R nerd, I decided to reproduce some of the findings using R. Obviously, the substance and ideas here are 100% attributable to AQR - my goal is add to our R volatility toolbox

(If you want to start at the beginning with three posts on portfolio volatility, see this intro post and this post on rolling vol and this post on visualization.) 
 

```{r}
symbols <- c("^GSPC", "^VIX")

prices <- 
  getSymbols(symbols, src = 'yahoo', from = "1990-01-01", 
             auto.assign = TRUE, warnings = FALSE) %>% 
  map(~Cl(get(.)))
  
# get daily, cont compounded returns
sp500_returns <- na.omit(ROC(GSPC$GSPC.Close, 1, type = "continuous"))
# spy_retcal_returns <- na.omit(CalculateReturns(spy_close, method = "compound"))
# spy_perret_returns <- na.omit(periodReturn(spy_close, period = 'daily', type = 'log'))
  
  
sp500_rolling_sd_20 <- rollapply(sp500_returns,
                             20,
                             function(x) StdDev(x))

sp500_rolling_sd_60 <- rollapply(sp500_returns,
                             60,
                             function(x) StdDev(x))
  
sp500_rolling_sd_annualized_20 <- (round((sqrt(252) * sp500_rolling_sd_20 * 100), 2))
sp500_rolling_sd_annualized_60 <- (round((sqrt(252) * sp500_rolling_sd_60 * 100), 2))

vol_vix_xts <- merge.xts(VIX$VIX.Close, sp500_rolling_sd_annualized_20, sp500_rolling_sd_annualized_60)

vol_vix_df <- 
  vol_vix_xts %>% 
  as_tibble(preserve_row_names = TRUE) %>% 
  mutate(date = ymd(row.names)) %>% 
  select(date, everything(), -row.names) %>% 
  rename(vix = VIX.Close, realized_vol_20 = GSPC.Close, realized_vol_60 = GSPC.Close.1) 

```


Alright, let's move on to consider what the VIX might be telling us about the future.

```{r, message = FALSE, warning = FALSE}

start.date <- index(sp500_rolling_sd_annualized_20[1])

end.date <- index(sp500_rolling_sd_annualized_20[nrow(sp500_rolling_sd_annualized_20) - 19])

# Extract the data, not the time index since we're adding to a tibble.
future_sp500_vol <- as_tibble(coredata(sp500_rolling_sd_annualized_20[20:nrow(sp500_rolling_sd_annualized_20), ]))

vix_sp500_future_vol <- 
  VIX$VIX.Close[paste(start.date, end.date, sep = "::")] %>% 
  as_tibble(preserve_row_names = TRUE) %>% 
  mutate(date = ymd(row.names)) %>% 
  select(-row.names) %>%
  na.omit() %>% 
  select(date, everything()) %>% 
  bind_cols(future_sp500_vol[1]) %>% 
  rename(vix = VIX.Close, Spy.future = GSPC.Close) 
 
tail(vix_sp500_future_vol)
```

```{r}
line_future_vol20 <- 
  # Plot vix on the x axis and future vol on the y axis
  ggplot(vix_sp500_future_vol, aes(date)) + 
  geom_line(aes(y = Spy.future, colour = "SP500")) +
  geom_line(aes(y = vix, colour = "Vix")) +
  scale_color_manual(values = c(SP500 = 'blue', Vix = 'red')) +
  # The remaining lines are aesthetic, title, axis labels.
  ggtitle("Vix and SP500 Subsequent Volatility") +
  xlab("date") +
  ylab("percent") +
  scale_y_continuous(labels = function(x){ paste0(x, "%") }, 
                     breaks = scales::pretty_breaks(n=10)) +
  theme(plot.title = element_text(hjust = 0.5))

line_future_vol20
```




```{r, message=FALSE, warning=FALSE}
# Assign different colors and shapes for obesrvations since the election.
# I'll go with pink diamonds for observations since the election.
# There might be a better way to assign colors and shapes conditional on dates but
# I like the simplicity here..
vix_sp500_future_vol$color <- ifelse(vix_sp500_future_vol$date <= "2016-11-08", 'light blue', 'pink')
vix_sp500_future_vol$shape <- ifelse(vix_sp500_future_vol$date <= "2016-11-08", 15, 18)

scatter_future_vol20 <- 
  # Plot vix on the x axis and future vol on the y axis
  ggplot(vix_sp500_future_vol, aes(vix, Spy.future)) + 
  geom_point(aes(vix, Spy.future), 
             colour = vix_sp500_future_vol$color, 
             shape = vix_sp500_future_vol$shape) +
  # Overlay a line of vix points, to make it easier to see what vix was predicting
  # on each each day.
  geom_line(aes(vix, vix)) +
  # The remaining lines are aesthetic, title, axis labels.
  ggtitle("Vix v. Subsequent Volatility") +
  xlab("Vix") +
  ylab("Realized vol 20 subsequent trading days") +
  scale_y_continuous(labels = function(x){ paste0(x, "%") }, 
                     breaks = scales::pretty_breaks(n=10)) +
  scale_x_continuous(labels = function(x){ paste0(x, "%") },
                     breaks = scales::pretty_breaks(n=20)) +
  theme(plot.title = element_text(hjust = 0.5))


scatter_future_vol20
```


For well-articulated insights and theory, have a look at the original [piece on Bloomberg](https://www.bloomberg.com/view/articles/2017-07-03/what-history-says-about-low-volatility). Briefly, the pink diamonds indicate we have been in a low volatility regime since the election, both measured by VIX and realized volatility. Another interesting note is that since 1990, we have never seen subsequent one-month volatiliy reach 20% when the VIX was below 12% (add a vertical line at 12% to see this more clearl). Said another way, spikes in realized volatility tend to follow spikes in the VIX.   

That's all for today. The main goal was to reproduce some interesting visualizations and insights about the VIX and realized volatility and hopefully that can serve as a starting point for future analysis using R. Thanks for reading and next time we'll wrap this into a Shiny app.